---
title: "STA 138 Discussion 6 -- solutions"
author: "Fall 2020"
header-includes:
   - \usepackage{enumerate,graphicx,nicefrac,amsmath}
   - \DeclareMathOperator*{\argmax}{arg\,max}
geometry: margin=0.5in
output: pdf_document
---


\section*{Likelihood ratio tests for multinomial models}
\emph{For our discussion this week, we will implement likelihood ratio tests. To do so, we'll pic up where we left off last week.}


Suppose that Amelia has sampled 76 newts out of a tank. There are four species in the tank (A, B, C, and D); the observed counts are given in the table below.



\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
  \hline
  A & B & C & D \\ 
  \hline
  14 & 22 & 25 & 15 \\ 
   \hline
\end{tabular}
\end{table}

Beatrice, the lab assistant, feeds the newts in the tank regularly. She takes particular notice when she does so of the brightly colored species A and B. From her experience, she claims that 20% of the newts in the tank are from species A, and 30% from species B. 

Assuming that she is correct, use numerical methods to obtain the MLE of the proportions of the species in the tank.

In the problems below, we will interpret the test results individually; there's no need to control for multiple testing. We'll carry out each test at $\alpha=0.01$.

\begin{enumerate}
\item Let's test whether we can rule out the possibility that the four species in the tank exist in equal proportions. Carry out a likelihood ratio test of
\[H_0:\pi_1=\pi_2=\pi_3=\pi_4=0.25\] against \[H_1:\pi_i \ne \pi_j \text{ for some } i \neq j\]

Remember that the likelihood ratio test asks us to compare sets of possible parameters $\Theta_0$ and $\Theta$, where the former represents the null hypothesis. In this case, let's write

\[\Theta_0 = \{(\pi_1,\pi_2,\pi_3,\pi_4): \pi_1=\pi_2=\pi_3=\pi_4=0.25 \}\]

and

\[\Theta = \left\{(\pi_1,\pi_2,\pi_3,\pi_4): 0 \le \pi_1,\pi_2,\pi_3,\pi_4 \le 1, \sum_{j=1}^4 \pi_j = 1\right\}\]

```{r,echo=FALSE}
## problem 1
y <-  c(14, 22, 25, 15)
lik0 <- dmultinom(y,prob=rep(0.25,4))
lik1 <- 0.00160
lrtstat <- -2*log(lik0/lik1)
pval <- 1-pchisq(lrtstat,3)
```

We can find without difficulty that the maximum likelihood under $H_0$ is
\[\argmax_{\pi \in \Theta_0} L(\pi,y) = L(\pi_0,y) \approx `r format(round(lik0,5),scientific=FALSE)`\ .\]

Furthermore, we found last week that
\[\argmax_{\pi \in \Theta} L(\pi,y) \approx 0.00160\ .\]

Is the difference between these significant at $\alpha=0.01$? From the above, we compute the LRT statistic to be `r round(lrtstat,5)`. Because $\Theta$ has $\dim(\Theta)=3$ freely varying parameters, and $\Theta_0$ has none, we use this in combination with a $\chi^2$ distribution with $3-0=3$ degrees of freedom to obtain $p-$value `r round(pval,3)`.

So, at $\alpha=0.01$, we fail to reject $H_0$.

\item Let's put Beatrice's claim to the test: using a likelihood ratio test, test \[H_0:\pi_1=0.2,\pi_2=0.3\] against \[H_1:\pi_1 \ne 0.2 \text{ or } \pi_2\ne 0.3\]

```{r,echo=FALSE}
## problem 2
lik0 <- 0.00143
lrtstat <- -2*log(lik0/lik1)
pval <- 1-pchisq(lrtstat,2)
```


Here $\Theta_1$ is as it was in the previous problem, while \[\Theta_0 = \{(\pi_1,\pi_2,\pi_3,\pi_4): \pi_1=0.2,\pi_2=0.3,0\le \pi_3,\pi_4\le1, \pi_3+\pi_4=0.5 \}\]

Since $\Theta_0$ now has one freely varying parameter, we have $\dim(\Theta_0)=1$.

Last week, we found that 
\[\argmax_{\pi \in \Theta_0} L(\pi,y) \approx 0.00143\ .\]

This gives us LRT statistic `r round(lrtstat,5)`, and, from the $\chi^2_{3-1}$ null distribution, we get $p-$value `r round(pval,3)`. So, we fail to reject this this claim as well.




\item \emph{Assuming that Beatrice's claim is true}, let's test whether there is evidence to rule out the possibility that species C and D occur in the tank in equal proportions. \[H_0:\pi_1=0.2,\pi_2=0.3,\pi_3=\pi_4=0.25\] against \[H_1: \pi_1=0.2 \text{ and }\pi_2=0.3 \text{ but either } \pi_3 \ne 0.25 \text{ or } \pi_4\ne 0.25\]

```{r,echo=FALSE}
## problem 3
lik1 <- lik0
lik0 <- dmultinom(y,prob=c(0.2,0.3,0.25,0.25))
lrtstat <- -2*log(lik0/lik1)
pval <- 1-pchisq(lrtstat,1)
```

Here, we have \[\Theta = \{(\pi_1,\pi_2,\pi_3,\pi_4): \pi_1=0.2,\pi_2=0.3,0\le \pi_3,\pi_4\le1, \pi_3+\pi_4=0.5 \}\]

and \[\Theta_0 = \{(\pi_1,\pi_2,\pi_3,\pi_4): \pi_1=0.2, 0.3, \pi_3=\pi_4=0.25 \}\]

This gives us LRT statistic `r round(lrtstat,5)`, and, from the $\chi^2_{1-0}$ null distribution, we get $p-$value `r round(pval,3)`. So, we fail to reject this this claim too!


\end{enumerate}


\subsection*{Appendix: R Script}

```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
