---
title: "STA 138 Discussion 1 Solutions"
author: "Fall 2020"
header-includes:
   - \usepackage{enumerate,graphicx}
geometry: margin=0.5in
output: pdf_document
---

```{r, echo = FALSE,warning = FALSE, message = FALSE}
library(ggplot2)
patients101 = read.csv("/home/andfarris/138-Fa20/labs/discussion1/patients101.csv")
```

1. (Note that solutions here illustrate both base R and ggplots; use of either or both is fine!)

(a) 

```{r, echo = FALSE,eval = TRUE}
#Problem 1 (a)
hist(patients101$sysBP, breaks=10, xlab = "Systolic Blood Pressure",main = "Distribution of Systolic Blood 
     Pressure 1(a)")
#ggplot(patients101, aes(x = sysBP)) + geom_histogram(binwidth = 5,color = "black",fill = "white")+
#  xlab("Systolic Blood Pressure")+ggtitle("1(a): Distribution of systolic blood pressure")
```

(b) With the exception of a single relative outlier, the data appears fairly symmetric.

\pagebreak

(c) 


```{r, echo = FALSE,eval = TRUE}
#Problem 1 (c)
boxplot(patients101$sysBP, horizontal = TRUE,main = "Distribution of Systolic Blood Pressure 1(c)")
#ggplot(patients101, aes(y=sysBP, x = factor("")))+ geom_boxplot() + ylab("Systolic Blood Pressure") + xlab(" ")+ coord_flip() + ggtitle("1(a): Distribution of systolic blood pressure")
```

(d) There is one clear, unusually large observation apparent from the boxplot.  

\newpage

2.

(a)

```{r, echo = FALSE,eval = TRUE}
#Problem 2 (a)
plot(patients101$weight, patients101$sysBP, main = "2(a):Relationship between Systolic Blood 
     Pressure and Weight",xlab = "Weight in pounds",ylab = "Systolic BP",pch = 19)
#qplot(weight, sysBP, data = patients101) + ggtitle("2(a):Relationship between Systolic Blood Pressure and Weight") + xlab("Weight in pounds") + ylab("Systolic BP")
```

(b) There seems to be a slight positive trend; as weight increases, the average systolic blood pressure appears to increase slightly as well.

(c) 

```{r, echo = FALSE,eval = TRUE}
#Problem 2 (c)
plot(patients101$sedmins, patients101$sysBP, main = "2(c):Relationship between Systolic Blood Pressure and Sedentary Minutes",xlab = "Sedentary Minutes",ylab = "Systolic BP",pch = 19)
#qplot(sedmins, sysBP, data = patients101) + ggtitle("2(c):Relationship between Systolic Blood Pressure and Sedentary Minutes") + xlab("Sedentary Minutes") + ylab("Systolic BP")
```

(d) There does not seem to be a notable trend here.


\pagebreak

3. 

(a) 

```{r, echo = FALSE,eval = TRUE}
#Problem 3 (a)
#boxplot(sysBP ~ gender, data = patients101, main = "Blood Pressure by Gender",horizontal = TRUE)
ggplot(patients101,aes(y = sysBP,x =gender)) + geom_boxplot() + ylab("Systolic Blood Pressure") + xlab("Gender") + ggtitle("Blood Pressure by Gender") + coord_flip() 
```


(b) The histogram for males and females appear to completely overlap, so they show no significant difference.

(c)


```{r, echo = FALSE, fig.height = 3, fig.width = 3.75, fig.align = "center", eval = TRUE}
#Problem 3 (c)
ggplot(patients101, aes(x = sysBP)) + geom_histogram(binwidth = 10,color = "black",fill = "white") + facet_grid(obese ~.) +ggtitle("Systolic Blood Pressure by Obesity")
```

(d) The 'Normal' weight category appears to have lower systolic blood pressures on average than 'obese' and 'overweight,' but there does not appear to be much of a difference between 'overweight' and 'obese.'


\pagebreak

4. 

```{r, echo = FALSE}
#Problem 4
#part(a)
aveBP = round(mean(patients101$sysBP),4)
#part(b)
all.ave = aggregate(sysBP ~ marriage, data = patients101, mean)
#part(c)
all.sd = aggregate(sysBP ~ marriage, data = patients101, sd)
#part(d)
all.n= aggregate(sysBP ~ marriage, data = patients101, length)
#part(e)
fns.weight = fivenum(patients101$weight)
```

(a) The average systolic blood pressure is: `r aveBP`.

(b) A table of results follows:

\begin{table}[ht!]
\centering
\begin{tabular}{rlr}
  \hline
 & marriage & ave sysBP \\ 
  \hline
1 & divorced & 133.50 \\ 
  2 & married & 124.77 \\ 
  3 & nevermarried & 115.50 \\ 
  4 & other & 126.00 \\ 
  5 & widowed & 132.75 \\ 
   \hline
\end{tabular}
\end{table}

(c) A table of results follows:

\begin{table}[ht!]
\centering
\begin{tabular}{rlr}
  \hline
 & marriage & sd sysBP \\ 
  \hline
1 & divorced & 22.85 \\ 
  2 & married & 21.34 \\ 
  3 & nevermarried & 19.27 \\ 
  4 & other & 22.21 \\ 
  5 & widowed & 8.81 \\ 
   \hline
\end{tabular}
\end{table}
(d) A table of results follows: 
\begin{table}[ht!]
\centering
\begin{tabular}{rlr}
  \hline
 & marriage & number \\ 
  \hline
1 & divorced &  12 \\ 
  2 & married &  52 \\ 
  3 & nevermarried &  16 \\ 
  4 & other &  12 \\ 
  5 & widowed &   8 \\ 
   \hline
\end{tabular}
\end{table}

(e) The five number summary is: `r fns.weight[1]` (min), `r fns.weight[2]` (Q2), `r fns.weight[3]` (Median), `r fns.weight[4]` (Q3), `r fns.weight[5]` (Max)


\hrulefill
\clearpage
\begin{center} Code Appendix \end{center}
\begin{verbatim}
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
\end{verbatim}
\end{enumerate}