---
title: "STA 138 Discussion 2 Solutions"
author: "Fall 2020"
header-includes:
   - \usepackage{enumerate,graphicx}
geometry: margin=0.5in
output: pdf_document
---


\begin{enumerate}
%Problem 1%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
```{r,echo=FALSE}
## 1.(a)
p1a <- pbinom(30,120,0.2)
## 1.(b)
p1b <- 1-pbinom(24-1,120,0.2)
## 1.(c)
p1c <- 1-pbinom(25-1,120,0.2)
```
\item Suppose that $Y \sim \text{Binomial}(120,0.2)$. 
\begin{enumerate}[(a)]
\item $P(Y \le 30) \approx `r round(p1a,4)`$ 
\item $P(Y \ge 24) \approx `r round(p1b,4)`$
\item $P(Y \ge 25) \approx `r round(p1c,4)`$
\end{enumerate}
%Problem 2%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
```{r,echo=FALSE}
## 2.(a)
p2a <- dhyper(4,20,80-20,20)
## 2.(b)
p2b <- phyper(3,20,80-20,20)
## 2.(c)
p2c <- phyper(10,20,80-20,20)
## 2.(d)
p2d <- phyper(6,20,80-20,20) - phyper(3,20,80-20,20)
```
\item Suppose that $Y \sim \text{Hypergeometric}(80,20,20)$
\begin{enumerate}[(a)]
\item $P(Y = 4) \approx `r round(p2a,4)`$
\item $P(Y \le 3) \approx `r round(p2b,4)`$
\item $P(Y \le 10) \approx `r round(p2c,4)`$
\item $P(4 \le Y \le 6) \approx `r round(p2d,4)`$
\end{enumerate}
%Problem 3%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
```{r,echo=FALSE}
## 3.(a)
p3a <- 1-ppois(8,9)
## 3.(b)
p3b <- ppois(15,9)-ppois(2,9)
```
\item Suppose that $Y \sim \text{Poisson}(9)$
\begin{enumerate}[(a)]
\item $P(Y > 8) \approx `r round(p3a,4)`$
\item $P(3 \le Y \le 15) \approx `r round(p3b,4)`$
\end{enumerate}
%Problem 4%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
```{r,echo=FALSE}
## 4.(a)
p4a <- dbinom(10,10,0.1)
## 4.(b)
p4b <- 10*dbinom(10, 10, 0.1)
## 4.(c)
p4c <- dmultinom(rep(1,10), prob=rep(0.1,10))
## 4.(d)
p4d <- (dmultinom(rep(1,10), prob=rep(0.1,10)) 
        + dmultinom(c(rep(1,8),2,0), prob=rep(0.1,10)) 
        + dmultinom(c(rep(1,8),0,2), prob=rep(0.1,10)))
```
\item Suppose that $(Y_1,Y_2,\ldots,Y_{10}) \sim \text{Multinomial}(10,0.1,0.1,\ldots,0.1)$. % note: Multinomial(n,\pi_1,\ldots,\pi_c)
\begin{enumerate}[(a)]
\item The marginal distribution of $Y_1$ is Binomial$(10,0.1)$, so using this we find $P(Y_1=10) = (0.1)^{10} \approx `r round(p4a,4)`$
\item Dividing this event into $10$ mutually exclusive events, where the $i$-th, $i=1, \cdots, 10$ event is that all $10$ subjects are assigned to the $i$-th category, we get $\approx `r format(round(p4b,5),scientific=FALSE)`$
\item Using the multinomial assignment of probabilities, we get $\approx `r format(round(p4c,5),scientific=FALSE)`$
\item Dividing this event up into the three resulting possible assignments of the two remaining subjects to the categories 9 and 10, and using the multinomial assignment of probabilities, we get $\approx `r format(round(p4d,5),scientific=FALSE)`$
\end{enumerate}
\end{enumerate}


\clearpage
\begin{center} Code Appendix \end{center}
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
