---
title: "STA 138 Discussion 7"
author: "Fall 2020"
header-includes:
   - \usepackage{enumerate,graphicx,nicefrac,amsmath}
   - \DeclareMathOperator*{\argmax}{arg\,max}
geometry: margin=0.5in
output: pdf_document
---

\begin{enumerate}
\item 

\begin{table}[ht]
\centering
\begin{tabular}{lcc}
  \hline
 & $C$ & $D$ \\ 
  \hline
$A$ & 16 &  110 \\ 
$B$   &  4 &  20 \\ 
   \hline
\end{tabular}
\end{table}

In the table above, you can find the results of an experiment by Kahneman and Tversky on peoples' risk preferences. The experiment consisted of eliciting choices by people for two decisions:

\begin{enumerate}[(i)]
\item Choose between:
\begin{enumerate}
\item[A.] a sure gain of \$240
\item[B.] 25\% chance for \$1000 gain, 75\% chance for \$0 gain
\end{enumerate}
\item Choose between:
\begin{enumerate}
\item[C.] a sure loss of \$750
\item[D.] 75\% chance to lose \$1000, 25\% chance to lose \$0
\end{enumerate}
\end{enumerate}

When people choose $B$ over $A$, we take them to be "risk loving" in choice (i); similarly, we take people who choose $D$ over $C$ to be "risk loving" in choice (ii). For the moment you can assume that our large sample approximation holds for this contingency table.

\begin{enumerate}
\item What do you estimate to be the odds ratio of choosing $B$ for $D$ vs. $C$? What does this tell you about risk preferences between the two choices?


\item Can you, using Pearson's $\chi^2$ test of independence, conclude that risk lovingness for (i) has an "effect" on risk livingess for (ii), at significance level $\alpha=0.1$?

```{r,echo=FALSE}
O <- matrix(c(16,110,
               4,20),
             byrow=TRUE,
             ncol=2) # observed
E <- rowSums(O)%*%t(colSums(O))/sum(O) # expected
pearsonStatistic <- sum((O-E)^2/E)
pearsonpVal <- 1-pchisq(pearsonStatistic,1) # df = 1 !
LRstatistic <- -2*sum(O*log(E/O))
LRpVal <- 1-pchisq(LRstatistic,1)

```



\item Can you, using a likelihood ratio test, conclude that risk lovingness for (i) has an "effect" on risk livingess for (ii), at significance level $\alpha=0.1$? 


\end{enumerate}


\item 

```{r,echo=FALSE}
counts <- matrix(c(79, 11, 
                   222, 19, 
                   169, 7, 
                   97, 8, 
                   107, 11, 
                   134, 9, 
                   133, 7, 
                   151, 16),
                 nrow=2)
timePeriods <- c("<2000 B.C.",
                      "2000-500 B.C.",
                      "500 B.C.-500 A.D.", 
                      "500-1200 A.D.",
                      "1200-1500 A.D.",
                      "1500-1700 A.D.",
                      "1700-1850 A.D.", 
                      ">1850 A.D.")
colnames(counts) <- c(as.roman(1:8))
rownames(counts) <- c("right","left")
O <- counts # observed
E <- rowSums(O)%*%t(colSums(O))/sum(O) # expected
pearsonStatistic <- sum((O-E)^2/E)
pearsonpVal <- 1-pchisq(pearsonStatistic,7) # df = 7 !
LRstatistic <- -2*sum(O*log(E/O))
LRpVal <- 1-pchisq(LRstatistic,7)
```

\begin{table}[ht]
\centering
\begin{tabular}{clllll}
  \hline
 & I: & $<$2000 B.C. &   & V: & 1200-1500 A.D. \\ 
   & II: & 2000-500 B.C. &   & VI: & 1500-1700 A.D. \\ 
   & III: & 500 B.C.-500 A.D. &   & VII: & 1700-1850 A.D. \\ 
   & IV: & 500-1200 A.D. &   & VIII: & $>$1850 A.D. \\ 
   \hline
\end{tabular}
\caption{Time periods}
\end{table}


\begin{table}[ht]
\centering
\begin{tabular}{lcccccccc}
  \hline
 & I & II & III & IV & V & VI & VII & VIII \\ 
  \hline
right &  79 & 222 & 169 &  97 & 107 & 134 & 133 & 151 \\ 
  left &  11 &  19 &   7 &   8 &  11 &   9 &   7 &  16 \\ 
   \hline
\end{tabular}
\caption{Contingency table}
\end{table}

A study counted the numbers of depictions of right- and left- handed people in artwork (Table 2) over different time periods (Table 1). Assuming that the artworks were randomly sampled, can you conclude that the relative frequencies of left- and right- handedness change over time?


\end{enumerate}

